\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2021-03-16 paper.tex
2 versions of \Lambda_ik

In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{equation}
  \label{eq:Lambda_ik-def}
  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} = \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i,
\end{equation}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{Construction of the samples in $S_i^{(w)}$}
\label{alg:construct-set-S}
\KwInput{$S_i^{(w)} = \varnothing$}
\KwOutput{$S_i^{(w)}$ filled with $2N$ samples}
Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i^{-1}$ \tcp{Cholesky decomposition} \nonumber
\For{$\breve{n} = 1,\ldots,N$}    
{ 
  Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
  Compute $\breve{b}_{\breve{n}} = C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
}
\textbf{Return:} {$S_i^{(w)}$}
\end{algorithm}

@@

$\text{M}_i= V_i^\top{\Sigma_i^{(w)}}^{-1}V_i + {D^{(w)}}^{-1}$ or $\text{M}_i= D^{(w)}$

\texttt{@VT: weird that $D^{(w)}$ and ${D^{(w)}}^{-1}$ ?}

@@

Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
  & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}). Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{Proximal gradient descent ISTA$_{\text{sg} l_1}$($v$, $f$, $\tilde \eta$, $\zeta$)}
\label{alg:ISTA}
\KwInput{A vector $v$, a function $f$ (its gradient), hyper-parameters $\tilde \eta$ and $\zeta$}
\KwOutput{$\flat^{\breve{w}^{max}}$}
Initialize $\flat^{(0)}=v$ \\
\For{$\breve{w} = 0,\ldots,\breve{w}^{max}$}    
{ 
  \vspace{-.8cm}
  \begin{align*}
  \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{\breve{w}}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
  &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \Big)
  \end{align*}
  \vspace{-.3cm}
}
\textbf{Return:} {$\flat^{\breve{w}^{max}}$}
\end{algorithm}

\vspace{.5cm}
TODO w/ VT: note on $\breve{w}^{max}$. Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \nonumber \\ 
  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \nonumber \\
  &+ U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.

>>>>

In the case $g_i(b_i, G_i) = \tilde g_i(b_i)$, one has
\begin{equation}
  \label{eq:def-E-g-tilde}
  \E_{\theta^{(w)}}[ \tilde g_i(b_i) | t_i, \delta_i, y_i] = \int_{\R^r} \tilde g_i(b_i) f(b_i | t_i, \delta_i, y_i ; \theta^{(w)}) \dd b_i = \dfrac{ \sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,\tilde g_i}^{\theta^{(w)}}}{\sum_{k=0}^{K-1} \pi_{\xi_k^{(w)}}(x_i) \Lambda_{ik,1}^{\theta^{(w)}}}
\end{equation}
with
\begin{align}
  \label{eq:Lambda_ik-b_i|y_i}
  \Lambda_{ik,\tilde g_i}^{\theta^{(w)}} &= \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i, y_i | b_i, G_i = k ; \theta^{(w)}) f(b_i ; \theta^{(w)}) \dd b_i \\
  \label{eq:Lambda_ik-b_i}
  &= f(y_i | G_i = k) \int_{\R^r} \tilde g_i(b_i) f(t_i, \delta_i | b_i, G_i = k ; \theta^{(w)}) f(b_i | y_i, G_i = k; \theta^{(w)}) \dd b_i
\end{align}
and $\Lambda_{ik,1}^{\theta^{(w)}}$ obtained from $\Lambda_{ik,\tilde g_i}^{\theta^{(w)}}$ taking $\tilde g_i: b_i \mapsto 1$.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{Construction of the samples in $S_i^{(w)}$}
\label{alg:construct-set-S}
\KwInput{$S_i^{(w)} = \varnothing$}
\KwOutput{$S_i^{(w)}$ filled with $2N$ samples}
Compute $C_i^{(w)} \in \R^{r \times r}$ such that 
$C_i^{(w)}{C_i^{(w)}}^\top = \text{M}_i$ \tcp{Cholesky decomposition} \nonumber
\For{$\breve{n} = 1,\ldots,N$}    
{ 
  Sample $\Omega_{\breve{n}} \sim \cN(0, I_r)$ \\
  Compute $\breve{b}_{\breve{n}} = C_i^{(w)} \Omega_{\breve{n}} \in \R^r$ \\
  Update $S_i^{(w)} \leftarrow S_i^{(w)} \cup \{\breve{b}_{\breve{n}}, -\breve{b}_{\breve{n}} \}$
}
\textbf{Return:} {$S_i^{(w)}$}
\end{algorithm}

@@

$\text{M}_i= W_i$ or $\text{M}_i= D^{(w)}$

@@

Now, for the update of $\beta_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  R^{(w)}_{n,k}(\beta_k) = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ &\delta_i \hat \E_{\theta^{(w)}}[ \tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2(b_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \\
  & + (y_i \odot \Phi_i^{(w)})^\top \hat \E_{\theta^{(w)}}[ M_{ik} | t_i, \delta_i, y_i] \\
  & - \hat \E_{\theta^{(w)}}[ c_{\phi^{(w)}}(M_{ik}) | t_i, \delta_i, y_i] \Big] 
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ depending on $\beta_k$, with $\tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1$ and $\tilde g_{t_i, \gamma_k^{(w)}, \beta_k}^2$ defined in~\eqref{eq:g-tilde-1-2}. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1, \ldots, t_J)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times.
The update for $\beta_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-beta_k}
  \beta_k^{(w+1)} \in \argmin_{\beta_k \in \R^q} R^{(w)}_{n,k}(\beta_k) + \zeta_{2,k} \norm{\beta_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}
The previous trick we used to handle the non-differentiability of the $\ell_1$ part in the elastic net is no longer sufficient for~\eqref{eq:minimization-beta_k}, since the $\ell_2$ norm in the sparse group lasso is not squared, leading to a non-differentiability at 0 (which is precisely the criterion exploited by the penalty to set some groups of coefficients to exactly 0~\citep{simon2013sparse}). Then, we choose to solve problem~\eqref{eq:minimization-beta_k} using the proximal gradient descent~\citep{boyd2004convex} described in Algorithm~\ref{alg:ISTA}.

@@

\begin{algorithm}[H]
\DontPrintSemicolon
\caption{Proximal gradient descent ISTA$_{\text{sg} l_1}$($v$, $f$, $\tilde \eta$, $\zeta$)}
\label{alg:ISTA}
\KwInput{A vector $v$, a function $f$ (its gradient), hyper-parameters $\tilde \eta$ and $\zeta$}
\KwOutput{$\flat^{\breve{w}^{max}}$}
Initialize $\flat^{(0)}=v$ \\
\For{$\breve{w} = 0,\ldots,\breve{w}^{max}$}    
{ 
  $\begin{aligned}
  \flat^{(\breve{w}+1)} \in &\argmin_u \Big\{\dfrac{1}{2\varrho_{\breve{w}}} \left\lVert u - \big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \big)\right\rVert_2^2 + \zeta \norm{u}_{\text{sg} l_1, \tilde{\eta}} \Big\} \\
  &\coloneqq\text{prox}_{\text{sg} l_1, \tilde{\eta}, \zeta} \Big(\flat^{(\breve{w})} - \varrho_{\breve{w}} \dfrac{\partial f(u)}{\partial u} \Big)
  \end{aligned}$
}
\textbf{Return:} {$\flat^{\breve{w}^{max}}$}
\end{algorithm}

\vspace{.5cm}
TODO w/ VT: note on $\breve{w}^{max}$ and $\varrho_{\breve{w}}$. Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\beta_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:beta-update}
  \beta_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}(\beta_k^{(w)}, R^{(w)}_{n,k}, \tilde \eta, \zeta_{2,k}).
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}.
The gradient of $R_{n,k}^{(w)}$ for the Gaussian case is here given by
\begin{align}
  \label{eq:grad-beta_k}
  \frac{\partial R_{n,k}^{(w)}(\beta_k)}{\partial \beta_k^l} = -n^{-1} &\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[\sum_{a=1}^\mathcal{A} {\gamma_{k,a}^l}^{(w)} \Big( \delta_i  \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_i, \beta_k^l, b_i^l)}{\partial \beta_k^l} | t_i, \delta_i, y_i] \nonumber \\ 
  &- \sum_{j=1}^J \lambda_0^{(w)}(t_j) \hat \E_{\theta^{(w)}}[ \frac{\partial \varphi_a(t_j, \beta_k^l, b_i^l)}{\partial \beta_k^l} \tilde g_{i, t_j, {\gamma_k^0}^{(w)}, \gamma_k^{(w)}, \beta_k}^1(b_i) | t_i, \delta_i, y_i] \ind{\{t_j\leq t_i\}} \Big) \nonumber \\
  &+ U_{il}^\top {\phi_l^{(w)}}^{-1}I_{n_{il}} (y_i^l - U_{il}\beta_k^l - V_{il}\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]
\end{align}
for all $l = 1, \ldots, L$. Note that the gradients of the considered $\varphi_a$ functions with respect to $\beta_k^l$ are given in Table~\ref{table:shared_associations}.
